Lead distance under a pickoff limit in Major League Baseball:\\A sequential game model

### 2025-12-24 articles/arxiv/paper.tex
clean up citations

\title{
  Lead distance under a pickoff limit in Major League Baseball:\\A sequential game model
}
\author[1]{Scott Powers}
\author[2]{Sivaramakrishnan Ramani}
\author[3]{Jacob Hahn}
\author[2]{Andrew J. Schaefer}
\affil[1]{Department of Sport Management, Rice University}
\affil[2]{Dept of Computational Applied Mathematics and Operations Research, Rice University}
\affil[3]{Department of Statistics, Rice University}

@@

    Section \ref{sec:sg-model} uses a two-payer zero-sum sequential game to model the evolving interaction between the runner and the pitcher.  Our model captures the vital aspects of the play while retaining analytical and computational tractability. Our model is closely related to the class of stochastic/dynamic Stackelberg games \parencite{goktas2022zero,li2017review,lopez2022stationary,vasal2020stochastic,vorobeychik2012computing}, which combines the elements of Stackelberg and stochastic games. The main difference between our infinite-horizon model and the stochastic Stackelberg games literature is that we do not discount the rewards. The rationale for this modeling choice is that our optimization objective---expected runs in the inning---by definition includes no discount. We establish a maximin Bellman equation for our sequential game and propose a variant of value iteration and policy iteration to determine the optimal strategies for runner and pitcher.

@@

    We note that fixing a pitcher policy reduces our two-player sequential game to a single-player MDP whose objective is to optimize the lead distance for the first-base runner. Optimality criterion similar to the two-player sequential game also exists for MDP with undiscounted rewards \parencite[Section 7.2]{bertsekas2012dynamic}. Numerical experiments using this decision  model for a specific pitcher policy are presented in Section \ref{sec:two-foot-rule}. These results may provide more immediately actionable insights for lead distance selection, as we discuss in Section \ref{sec:discussion}.

@@

       Under Assumption \ref{assum:game_halts}, we prove that the expected number of runs scored by the batter, $V^{\pi_\text{R},\pi_\text{P}}$, in  \eqref{eqn:sg_pol_val} is finite. Our proof is based on standard arguments used in the analyses of undiscounted problems \parencite[Section 7.2]{bertsekas2012dynamic}. We present it here for completeness.

@@

        We present optimality characterizations in Theorem \ref{lem:vi_convergence} and Propositions \ref{lem:bellman_equations} to  \ref{lem:policy_iteration}. These results and their proofs are similar to the undiscounted problem in the single-player case \parencite[Section 7.2]{bertsekas2012dynamic} and the zero-sum stochastic game with simultaneous moves \parencite{patek_stochastic_1999}. We state and prove our results for completeness and to outline the minor differences. 
      \begin{definition}\label{defn:sg_operators}
        \normalfont
        Consider the sequential game defined in Section \ref{sec:sg-model}. Fix a $u \in \mathbb{R}^{\abs{\cal S}}$ with $u(\Delta) = 0$.
        \begin{enumerate}
            \item For any $\pi_\text{R} \in \Pi_\text{R}$ and $\pi_\text{P} \in \Pi_\text{P}$, the evaluation operator $T_{\pi_\text{R},\pi_{\text{P}}}$ on $\mathbb{R}^{\abs{\cal S}}$ is defined as
            \begingroup
            \allowdisplaybreaks
            \begin{align}
            \label{eqn:eval_operator}
            (T_{\pi_\text{R},\pi_{\text{P}}} u)(s) &\defeq \mathbb{E}_{s' \sim p(\cdot|s,\pi_\text{R}(s),\pi_\text{P}(s,\pi_\text{R}(s)))}\left[r(s'|s) + u(s')\right] \ \forall s \in {\cal S}.
            \end{align}
            \endgroup
            \item The optimality operator $T$ on $\mathbb{R}^{\abs{\cal S}}$ is defined as
            \begingroup
            \allowdisplaybreaks
            \begin{align}
            \label{eqn:bellman_operator}
            (T u)(s) &\defeq \max_{a_\text{R} \in A_\text{R}(s)}\min_{a_\text{P} \in A_\text{P}(s,a_\text{R})}\mathbb{E}_{s' \sim p(\cdot|s,a_\text{R},a_\text{P})}\left[r(s'|s) + u(s')\right] \ \forall s \in {\cal S}.
            \end{align}
            \endgroup
        \end{enumerate}       
       \end{definition}

@@

           Fix a $s \in {\cal S}$. If $s = (b,c,o,d)$ is such that $b \neq (1,0,0)$, then from \eqref{eqn:runner_action_sapce} and \eqref{eqn:pitcher_action_sapce}  $A_\text{R}(s) = \{\delta\}$ and $ A_\text{P}(a_\text{R},s) = \{\delta\}$. As a result, $p(\cdot|s,a_\text{R},a_\text{P}) = p(\cdot|s)$, and hence the convergence is independent of $a_\text{R}$ and $a_\text{P}$. Suppose the component $b$ equals $(1,0,0)$. Let $\{a_\text{R}^k\} \in A_\text{R}(s)$ be such that $\lim_{k \to \infty}a_\text{R}^k = a_\text{R} \in A_\text{R}(s)$. Then from  \eqref{eqn:pitcher_action_sapce}, $A_\text{P}(a_\text{R}^k,s) = \{0,1\}$ for all $k \in \mathbb{N}$. Therefore
           \begingroup
           \allowdisplaybreaks
           \begin{align*}
           &\lim_{k \to \infty}\min_{a_\text{P} \in A_\text{P}(s,a_\text{R}^k)}\sum_{s' \in {\cal S}}\big[r(s'|s) + u_{k-1}(s')\big]p(s'|s,a_\text{R}^k,a_\text{P}) \\
           &= \lim_{k \to \infty}\min\Bigg(\sum_{s' \in {\cal S}}\big[r(s'|s) + u_{k-1}(s')\big]p(s'|s,a_\text{R}^k,0),\sum_{s' \in {\cal S}}\big[r(s'|s) + u_{k-1}(s')\big]p(s'|s,a_\text{R}^k,1)\Bigg)\\
           &= \min\Bigg(\sum_{s' \in {\cal S}}\big[r(s'|s) + V^*(s')\big]p(s'|s,a_\text{R},0),\sum_{s' \in {\cal S}}\big[r(s'|s) + V^*(s')\big]p(s'|s,a_\text{R},1)\Bigg)\\
           &= \min_{a_\text{P} \in A_\text{P}(s,a_\text{R}^k)}\sum_{s' \in {\cal S}}\big[r(s'|s) + V^*(s')\big]p(s'|s,a_\text{R},a_\text{P}),
           \end{align*}
           \endgroup
           where the second equality follows from Assumption \ref{assum:transition_pby_continuous} and the convergence of $u_{k-1}$ to $V^*$ established in Theorem \ref{lem:vi_convergence}. We have established the continuous convergence of the sequence of functions $\min_{a_\text{P} \in A_\text{P}(s,a_\text{R})}\sum_{s' \in {\cal S}}\big[r(s'|s) + u_{k-1}(s')\big]p(s'|s,a_\text{R},a_\text{P})$ over $A_\text{R}(s)$. The set $A_\text{R}(s)$ in \eqref{eqn:runner_action_sapce}   is compact. Hence, by Theorem 5 in Subsection X, Section 21 in  \parencite{kuratowski2014topology}, $\min_{a_\text{P} \in A_\text{P}(s,a_\text{R})}\sum_{s' \in {\cal S}}\big[r(s'|s) + u_{k-1}(s')\big]p(s'|s,a_\text{R},a_\text{P})$ converges to $\min_{a_\text{P} \in A_\text{P}(s,a_\text{R})}\sum_{s' \in {\cal S}}\big[r(s'|s) + V^*(s')\big]p(s'|s,a_\text{R},a_\text{P})$ uniformly over $A_\text{R}(s)$.

>>>>

\title{
  Lead distance under a pickoff limit in Major League Baseball:\\A sequential game model
}
\author[1]{Scott Powers}
\author[2]{Sivaramakrishnan Ramani}
\author[3]{Jacob Hahn}
\author[2]{Andrew J. Schaefer}
\affil[1]{Department of Sport Management, Rice University}
\affil[2]{Department of Computational Applied Mathematics and Operations Research, Rice University}
\affil[3]{Department of Statistics, Rice University}

@@

    Section \ref{sec:sg-model} uses a two-payer zero-sum sequential game to model the evolving interaction between the runner and the pitcher.  Our model captures the vital aspects of the play while retaining analytical and computational tractability. Our model is closely related to the class of stochastic/dynamic Stackelberg games \parencite{goktas_zero-sum_2022,li_review_2017,lopez_stationary_2023,vasal_sequential_2022,vorobeychik_computing_2012}, which combines the elements of Stackelberg and stochastic games. The main difference between our infinite-horizon model and the stochastic Stackelberg games literature is that we do not discount the rewards. The rationale for this modeling choice is that our optimization objective---expected runs in the inning---by definition includes no discount. We establish a maximin Bellman equation for our sequential game and propose a variant of value iteration and policy iteration to determine the optimal strategies for runner and pitcher.

@@

    We note that fixing a pitcher policy reduces our two-player sequential game to a single-player MDP whose objective is to optimize the lead distance for the first-base runner. Optimality criterion similar to the two-player sequential game also exists for MDP with undiscounted rewards \parencite[Section 7.2]{bertsekas_dynamic_2012}. Numerical experiments using this decision  model for a specific pitcher policy are presented in Section \ref{sec:two-foot-rule}. These results may provide more immediately actionable insights for lead distance selection, as we discuss in Section \ref{sec:discussion}.

@@

       Under Assumption \ref{assum:game_halts}, we prove that the expected number of runs scored by the batter, $V^{\pi_\text{R},\pi_\text{P}}$, in  \eqref{eqn:sg_pol_val} is finite. Our proof is based on standard arguments used in the analyses of undiscounted problems \parencite[Section 7.2]{bertsekas_dynamic_2012}. We present it here for completeness.

@@

        We present optimality characterizations in Theorem \ref{lem:vi_convergence} and Propositions \ref{lem:bellman_equations} to  \ref{lem:policy_iteration}. These results and their proofs are similar to the undiscounted problem in the single-player case \parencite[Section 7.2]{bertsekas_dynamic_2012} and the zero-sum stochastic game with simultaneous moves \parencite{patek_stochastic_1999}. We state and prove our results for completeness and to outline the minor differences. 
      \begin{definition}\label{defn:sg_operators}
        \normalfont
        Consider the sequential game defined in Section \ref{sec:sg-model}. Fix a $u \in \mathbb{R}^{\abs{\cal S}}$ with $u(\Delta) = 0$.
        \begin{enumerate}
            \item For any $\pi_\text{R} \in \Pi_\text{R}$ and $\pi_\text{P} \in \Pi_\text{P}$, the evaluation operator $T_{\pi_\text{R},\pi_{\text{P}}}$ on $\mathbb{R}^{\abs{\cal S}}$ is defined as
            \begingroup
            \allowdisplaybreaks
            \begin{align}
            \label{eqn:eval_operator}
            (T_{\pi_\text{R},\pi_{\text{P}}} u)(s) &\defeq \mathbb{E}_{s' \sim p(\cdot|s,\pi_\text{R}(s),\pi_\text{P}(s,\pi_\text{R}(s)))}\left[r(s'|s) + u(s')\right] \ \forall s \in {\cal S}.
            \end{align}
            \endgroup
            \item The optimality operator $T$ on $\mathbb{R}^{\abs{\cal S}}$ is defined as
            \begingroup
            \allowdisplaybreaks
            \begin{align}
            \label{eqn:bellman_operator}
            (T u)(s) &\defeq \max_{a_\text{R} \in A_\text{R}(s)}\min_{a_\text{P} \in A_\text{P}(s,a_\text{R})}\mathbb{E}_{s' \sim p(\cdot|s,a_\text{R},a_\text{P})}\left[r(s'|s) + u(s')\right] \ \forall s \in {\cal S}.
            \end{align}
            \endgroup
        \end{enumerate}       
       \end{definition}

@@

           Fix a $s \in {\cal S}$. If $s = (b,c,o,d)$ is such that $b \neq (1,0,0)$, then from \eqref{eqn:runner_action_sapce} and \eqref{eqn:pitcher_action_sapce}  $A_\text{R}(s) = \{\delta\}$ and $ A_\text{P}(a_\text{R},s) = \{\delta\}$. As a result, $p(\cdot|s,a_\text{R},a_\text{P}) = p(\cdot|s)$, and hence the convergence is independent of $a_\text{R}$ and $a_\text{P}$. Suppose the component $b$ equals $(1,0,0)$. Let $\{a_\text{R}^k\} \in A_\text{R}(s)$ be such that $\lim_{k \to \infty}a_\text{R}^k = a_\text{R} \in A_\text{R}(s)$. Then from  \eqref{eqn:pitcher_action_sapce}, $A_\text{P}(a_\text{R}^k,s) = \{0,1\}$ for all $k \in \mathbb{N}$. Therefore
           \begingroup
           \allowdisplaybreaks
           \begin{align*}
           &\lim_{k \to \infty}\min_{a_\text{P} \in A_\text{P}(s,a_\text{R}^k)}\sum_{s' \in {\cal S}}\big[r(s'|s) + u_{k-1}(s')\big]p(s'|s,a_\text{R}^k,a_\text{P}) \\
           &= \lim_{k \to \infty}\min\Bigg(\sum_{s' \in {\cal S}}\big[r(s'|s) + u_{k-1}(s')\big]p(s'|s,a_\text{R}^k,0),\sum_{s' \in {\cal S}}\big[r(s'|s) + u_{k-1}(s')\big]p(s'|s,a_\text{R}^k,1)\Bigg)\\
           &= \min\Bigg(\sum_{s' \in {\cal S}}\big[r(s'|s) + V^*(s')\big]p(s'|s,a_\text{R},0),\sum_{s' \in {\cal S}}\big[r(s'|s) + V^*(s')\big]p(s'|s,a_\text{R},1)\Bigg)\\
           &= \min_{a_\text{P} \in A_\text{P}(s,a_\text{R}^k)}\sum_{s' \in {\cal S}}\big[r(s'|s) + V^*(s')\big]p(s'|s,a_\text{R},a_\text{P}),
           \end{align*}
           \endgroup
           where the second equality follows from Assumption \ref{assum:transition_pby_continuous} and the convergence of $u_{k-1}$ to $V^*$ established in Theorem \ref{lem:vi_convergence}. We have established the continuous convergence of the sequence of functions $\min_{a_\text{P} \in A_\text{P}(s,a_\text{R})}\sum_{s' \in {\cal S}}\big[r(s'|s) + u_{k-1}(s')\big]p(s'|s,a_\text{R},a_\text{P})$ over $A_\text{R}(s)$. The set $A_\text{R}(s)$ in \eqref{eqn:runner_action_sapce}   is compact. Hence, by Theorem 5 in Subsection X, Section 21 in  \parencite{kuratowski_topology_2014}, $\min_{a_\text{P} \in A_\text{P}(s,a_\text{R})}\sum_{s' \in {\cal S}}\big[r(s'|s) + u_{k-1}(s')\big]p(s'|s,a_\text{R},a_\text{P})$ converges to $\min_{a_\text{P} \in A_\text{P}(s,a_\text{R})}\sum_{s' \in {\cal S}}\big[r(s'|s) + V^*(s')\big]p(s'|s,a_\text{R},a_\text{P})$ uniformly over $A_\text{R}(s)$.
